An in-depth study of Explainability and Adversarial Robustness for RNNs

### 2019-12-13 ex3/paper/paper.tex
Changed confidence plot
Changed sequence plots

As apparent from the above discussion, trying to explain how packet features influence the model's predictions, poses various difficulties. As a first step to explaining our model's predictions, we investigate how important features at different time steps are for the model's decisions. Intuitively, features at the beginning of a flow should be the most important while the model's predictions should not vary significantly anymore, as soon as the model has come to a decision.

\begin{figure}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_5_DoS - DDoS-DoS Hulk"}.pdf}
\includegraphics[width=\columnwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_6_DoS - DDoS-DoS Slowhttptest"}.pdf}
% \includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_full_no_ttl_pred_plots_outcomes_0_3.pickle_14_Web Attack-XSS"}.pdf}
% \includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_full_no_ttl_pred_plots_outcomes_0_3.pickle_11_PortScan-PortScan - Firewall off"}.pdf}

\caption{Median confidence per time step and first and third quartiles. For the majority of attacks types, confidence increases in the first few steps and then stays almost constant at 1.}
\label{fig:confidence}
\end{figure}

@@

@@

\caption{Exemplary PD plot and adversarial flows for the DoS Slowloris attack in CIC-IDS-2017. The solid line shows the mean value of the feature for all flows of the attack type. The shaded region shows the change in confidence that occurs when the feature is varied.}
%\todo{IAT plot is wrong: There shouldn't be step 0.} %AH. guess this is fine now
\label{fig:pred_plots2}
\end{figure}

Next, we want to investigate if our classifier just memorized its inputs. The most interesting packet features that can be used to visualize this in our use case are the \gls{iat} between packets in a flows and the packet length. To depict how the model comes to a decision, we determined the total minimum and maximum for these features that ever occur for all flows. For a given flow $X$ we then varied feature values at each time step between minimum and maximum and plot how the confidence changes when the feature varies (\autoref{fig:pred_plots2}) depict the averaged predictions for different attack families. It seems clear that the classifier didn't just memorize the inputs but actually learned them since there is always a smooth decrease of confidence if we change the feature values. Instead, when memorizing, we would expect a sharp decline of confidence to occur. Another interesting observation is that after the first couple of packets the classifier is certain about its decision and changing features doesn't change its confidence any longer. This allows the conclusion that if an attacker is able to change the first few packets, she could change the decision of the classifier. This is consistent with our observations in \autoref{sec:adv}. %We will further investigate this question in \autoref{sec:adv}.

>>>>

%As apparent from the above discussion, trying to explain how packet features influence the model's predictions, poses various difficulties. As a first step to explaining our model's predictions, we investigate how important features at different time steps are for the model's decisions.
Intuitively, features at the beginning of a flow should be the most important while the model's predictions should not vary significantly anymore, as soon as the model has come to a decision.

\begin{figure}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_5_DoS - DDoS-DoS Hulk"}.pdf}
\includegraphics[width=\columnwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_15_All samples"}.pdf}
% \includegraphics[width=\columnwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_6_DoS - DDoS-DoS Slowhttptest"}.pdf}
% \includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_full_no_ttl_pred_plots_outcomes_0_3.pickle_14_Web Attack-XSS"}.pdf}
% \includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_full_no_ttl_pred_plots_outcomes_0_3.pickle_11_PortScan-PortScan - Firewall off"}.pdf}

\caption{Classifier confidence per time step for CIC-IDS-2017. % and first and third quartiles.
For the majority of attack types, confidence increases in the first few steps and then stays almost constant at 1.}
\label{fig:confidence}
\end{figure}

@@

For investigating, how features influence the prediction outcome, we extend \glspl{pdp} to the sequential setting. 
Denoting as $X=\{\boldsymbol X^1, \ldots, \boldsymbol X^m\}$ the series of packet features $\boldsymbol X^t \in \mathbb R^n$ and $h_t(X)$ the network's hidden state at time step $t$, we define the sequential \gls{pdp} as 
\begin{align}
\text{PD}&\text{P}_{c,i}(t,w)= \\ &\mathbb E_{X | C} \Big(f \left(h_{t-1}( X), X_1^t,\ldots,X_{i-1}^t,w,X_{i+1}^t,\ldots X_n^t \right) | c\Big). \nonumber 
\end{align}

\autoref{fig:pred_plots2} shows an example together with the mean values for both unmodified samples and adversarial samples. Also in this figure we notice that mainly the first few packets are able to influence the prediction outcome and modifying features at a later time step does not change its confidence any longer. In many cases, the adversarial sample generation indeed moves packet features to areas where the network is less likely to be classified as attacks, confirming the effectiveness of \glspl{pdp}. In other cases, however, we did not observe an agreement between \gls{pdp} and adversarial modifications, hinting at dependencies which cannot be presented by \glspl{pdp}.

@@

\caption{Exemplary sequential PD plot and adversarial flows for the DoS Slowloris attack in CIC-IDS-2017. The solid line shows the mean value of the feature for all flows of the attack type. The shaded region shows the change in confidence that occurs when the feature is varied.}
%\todo{IAT plot is wrong: There shouldn't be step 0.} %AH. guess this is fine now
\label{fig:pred_plots2}
\end{figure}

%Next, we want to investigate if our classifier just memorized its inputs. The most interesting packet features that can be used to visualize this in our use case are the \gls{iat} between packets in a flows and the packet length. To depict how the model comes to a decision, we determined the total minimum and maximum for these features that ever occur for all flows. For a given flow $X$ we then varied feature values at each time step between minimum and maximum and plot how the confidence changes when the feature varies (\autoref{fig:pred_plots2}) depict the averaged predictions for different attack families.

Furthermore, it seems clear that the classifier did not just memorize the inputs but actually learned appropriate space representations, as there is always a smooth decrease of confidence when modifying feature values. When memorizing, we would  instead expect a sharp decline of confidence. % to occur. 
%Another interesting observation is that after the first couple of packets the classifier is certain about its decision and changing features doesn't change its confidence any longer. This allows the conclusion that if an attacker is able to change the first few packets, she could change the decision of the classifier. This is consistent with our observations in \autoref{sec:adv}. %We will further investigate this question in \autoref{sec:adv}.
